### 2016-03-10 paper/paper.tex
added final reward data to the paper
minor changes to the paper

\subsection{Actor Critic Relative Entropy Policy Search}
\emph{Actor critic relative entropy policy search} (AC-REPS) as introduced by
\cite{acreps} is a method for policy improvement based on \emph{relative entropy
policy search} (REPS) \cite{reps} which allows for model-free learning with
non-parametric continous policies and random exploration. It consists of the
following three steps:

@@

During the simulation we generate sample in the form $(s_i, a_i, r_i, s'_i)$,
where we start in state $s_i$, take action $a_i$ to get to state $s'_i$ and
get reward $r_i$. We also define a feature-based Q-function as $Q(s, a) =
\phi(s,a)^T \theta$ with feature function $\phi(s,a)$ and parameter vector
$\theta$. Details of our state and action representations and rewards and
features we used are discussed in section TODO.

@@

This choice of weights will reproduce actions wich lead to a higher reward with
higher probability but will also choose suboptimal actions with some
probability, because our new policy can not be too far away in terms of KL
divergence from the old policy and we start with a random policy. This achieves
a tradeoff between exploration and exploitation.

@@

Because this policy is relatively easy to learn and there are specialized path
planning algorithms like $A^*$ which can easily compute an optimal path through
any maze we did not learn this policy and instead focused on the second one,
which is much more interesting. Therefore we manually specified this policy by
providing a number of waypoints the object should pass and giving the next
waypoint as the target position. Once we reach a waypoint we move one to next
one.

@@

When requesting new samples from the simulator we specify the current policy
$\pi$ which will be used to generate these sampels as well as the sampling
parameters such as the shape of the object, the number of kilobots $\nkb$, the
number of episodes $\nep$ and the number of steps per episode $\nst$.

Given these parameters the simulator first generates $\nep$ kilobot start
positions $\mathit{start}_i$ equally spaced in a circle of radius $1.5 \cdot r$
around the object which starts in the middle of the screen where $r$ is the
radius or half width of the object depending on the its shape. Then for
$i \in \{1, \dots, \nep\}$ we do the following:

The object's position is reset to the middle of the screen and its angle is set
to 0. After that the light position is set to $\mathit{start}_i$ and the
position of the $\nkb$ kilobots is set to $\mathit{start}_i + \mathit{off}_j$,
where $\mathit{off}_j$ is a small individual offset for kilobot number $j$.
Then $\nst$ time steps are simulated where in each one an action $a_k$ is chosen
for the current state $s_k$ according to $\pi$ and applied by adding $a_k$ to
the current light positon. Next the kilobots move directly toward the light for
a short amount time (1 second for all of our simulations) and lastly we record
the current state as $s_k'$, compute the reward $r_k$ and save the tuple $(s_k,
a_k, r_k, s_k')$ as a sample. We also restrict the magnitude of the light and
kilobot movements to get a more realistic simulation.

@@

\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.9\linewidth]{figures/simulator_maze.png}
    \caption{Simulator showing 4 kilobots pushing an object through a maze}
    \label{fig:simulator}
\end{figure}

@@

For this policy we decided to only learn how to push the object to the right
instead of an arbitrary direction. This simplifies and speeds up the learning
process because we need to generate fewer samples and can represent our policy
using a smaller state subset for the sparse GP and therefore can compute the
policy update much faster. As discussed in the last paragraph of the previous
section we can still use this policy to push the object in an arbitrary
direction by rotation the state and action appropriately.

@@

We repeat these steps a fixed number of times and manually inspect the learned
policy. If it is not satisfactory (it can not solve the maze) we continue the
learing process.

@@

With $n$ kilobots our state tupel is
$$s = (l_x, l_y, k_{1x}, k_{1y}, \dots, k_{nx}, k_{ny}), $$
where $(l_x, l_y)$ is the light position and $(k_{ix}, k_{iy})$ is the position
of the $i-th$ kilobot. All positions are relative to the object's center
position.

Our actions tuple is
$$a = (a_x, a_y),$$
which is the direction the light should move in.

@@

Our reward is given by
$$ r = (c_x - p_x) - 0.5 \cdot \mathrm{abs} (c_y - p_y), $$
where $(c_x, c_y)$ is the object's current position and $(p_x, p_y)$ is the
object's position from the last time step. Therefore we reward movement to the
right and punish movement to the left and any movement in $y$ direction.

@@

For our learing process we need a feature function $\phi(s, a)$ over
state-action pairs to represent our Q-function, a feature function $\phi(s)$
over states for REPS and a kernel function $K(s_1, s_2)$ between states for the
GP.

@@

After that we compute the distance for the other dimensions (light positions for
$K$ and light positions and actions for $K'$) as
$$
d_{other} = (x_1 - x_2)^T \mathrm{diag}\left(\frac{1}{\sigma_{x}^2}\right) (x_1 - x_2),
$$
where $x_1$ and $x_2$ are vectors holding the values for all non-kilobot
dimensions and $\sigma_x$ is the non-kilobot bandwith.

@@

In our experiments we used $w = 0.5$ and recomputed $\sigma_{kb}$ and $\sigma_x$
in each iteration as the square root of the median of the squared distance
between all sample values for each dimension separately. We also scaled the
bandwidths uniformly in each dimension by some manually specified factors.

@@

For our final evaluation we made 20 runs of our learning process with 15
iterations each generating $25 \cdot 100 = 2500$ samples per iteration. Using
an Haswell Quad Core with 3.4GHz per core each iteration took about 3 minutes,
so the total runtime for each policy was about 45 minutes.

The policies start of only being able to push the object when the kilobots start
to the left of the object. After about 5 iterations they can also handle kilobot
starting positions above and below the object by first moving the kilobots to
the left and down and then pushing the object from the left side. After further
iterations the policies are able to push the object to the right even if the
kilobots start above and to the right of the object (see
\autoref{fig:policyEvolution}). With even more iterations the kilobot starting
positions from which the object can be pushed tend to extend even further, but
this improvement is inconsistent and the policies may even get worse again.
Therefore we manually inspected the policies at each iteration and choose a good
policy (which is typically found in the first about 15 iterations) for solving
the maze.

\begin{figure*}[!htb]
	\centering
	\subfloat[1 iterations]{
        \def\svgwidth{0.45\textwidth}
        \input{figures/policy_1.pdf_tex}
    }
    \hfill
	\subfloat[5 iterations]{
        \def\svgwidth{0.45\textwidth}
        \input{figures/policy_5.pdf_tex}
    }
    \hfill
	\subfloat[10 iterations]{
        \def\svgwidth{0.45\textwidth}
        \input{figures/policy_10.pdf_tex}
    }
    \hfill
	\subfloat[15 iterations]{
        \def\svgwidth{0.45\textwidth}
        \input{figures/policy_15.pdf_tex}
    }
    \hfill
	\caption[foo]{Evolution of the learned policy.}
	\label{fig:policyEvolution}
\end{figure*}

The value function also shows that kilobot positions left of the object are
preferrable to positions to the right and you can also somewhat see the value
function increasing when moving around the object from the right to the left
(see \autoref{fig:value}).

@@

\begin{figure}[!htb]
    \centering
    \resizebox{\linewidth}{!}{\input{figures/reward.tex}}
    \caption{Total reward for each iterations averaged over 8 runs.
             TODO: 20 runs are not complete yet}
    \label{fig:meanReward}
\end{figure}

We made the policy plots by getting the mean action for states distributed on a
grid and plotting it as an arrow indicating the direction the light should move
in. For this we set the relative light positions to evenly spaced values between
$-0.25$ and $0.25$ for the $x-$ and $y-$dimension and set the position for all
kilobots to coincide with the light position.

@@

To evaluate our policies we used our simulator to push the object through the
maze as described in \ref{sec:simulator}. We found that our learned policies are
sufficient to achieve this task and are even able to move the object when the
kilobots are (almost) completely on the wrong side which requires moving them
around the object. Moreover we can transfer our policy to more or less kilobots
and are still able to push the object because of our choice of kernel. We
learned the policy with 4 kilobots and testes it with 1,2,3,4,6,8 and 10
kilobots and it was able to push the object through the maze in each case.
Finally we were also able to push a circel through the maze even though the
policy was learned on a square.

@@

Although this policy works well in the simulator it is not directly applicable
to the real kilobots because our simulator is overly simplistic. It moves
the kilobots directly toward the light instead of using phototaxis and also
allows for arbitrary light movements where in reality the light would be moved
by a robot arm which has some constraints on the forces used to move it.

Therefore a future work could simulate the kilobots more realistically and try
to transfer the learned policy to the real kilobots by building a maze and
controlling a robot arm holding a flaslight usign the learned policy.

>>>>

\subsection{Actor Critic Relative Entropy Policy Search}
\emph{Actor critic relative entropy policy search} (AC-REPS) as introduced by
\cite{acreps} is a method for policy improvement based on \emph{relative entropy
policy search} (REPS) \cite{reps} which allows for model-free learning with
non-parametric continuous policies and random exploration. It consists of the
following three steps:

@@

During the simulation we generate sample in the form $(s_i, a_i, r_i, s'_i)$,
where we start in state $s_i$, take action $a_i$ to get to state $s'_i$ and
get reward $r_i$. We also define a feature-based Q-function as $Q(s, a) =
\phi(s,a)^T \theta$ with feature function $\phi(s,a)$ and parameter vector
$\theta$. Details of our state and action representations and rewards and
features we used are discussed in section \ref{sec:objpolicy}.

@@

This choice of weights will reproduce actions which lead to a higher reward with
higher probability but will also choose suboptimal actions with some
probability, because our new policy can not be too far away in terms of KL
divergence from the old policy and we start with a random policy. This achieves
a trade-off between exploration and exploitation.

@@

Because this policy is relatively easy to learn and there are specialized path
planning algorithms like $A^*$ which can easily compute an optimal path through
any maze we did not learn this policy and instead focused on the second one,
which is much more interesting. Therefore we manually specified this policy by
providing a number of waypoints the object should pass and give the next
waypoint as the target position. Once we reach a waypoint we move one to next
one.

@@

When requesting new samples from the simulator we specify the current policy
$\pi$ which will be used to generate these samples as well as the sampling
parameters such as the shape of the object, the number of kilobots $\nkb$, the
number of episodes $\nep$ and the number of steps per episode $\nst$.

Given these parameters the simulator first generates $\nep$ kilobot start
positions $\mathit{start}_i$ equally spaced in a circle of radius $1.5 \cdot r$
around the object which starts in the middle of the screen, where $r$ is the
radius or half-width of the object depending on the its shape. Then for
$i \in \{1, \dots, \nep\}$ we do the following:

The object's position is reset to the middle of the screen and its angle is set
to 0. After that the light position is set to $\mathit{start}_i$ and the
position of the $\nkb$ kilobots is set to $\mathit{start}_i + \mathit{off}_j$,
where $\mathit{off}_j$ is a small individual offset for kilobot number $j$.
Then $\nst$ time steps are simulated where in each one an action $a_k$ is chosen
for the current state $s_k$ according to $\pi$ and applied by adding $a_k$ to
the current light position. Next the kilobots move directly toward the light for
a short amount time (1 second for all of our simulations) and lastly we record
the current state as $s_k'$, compute the reward $r_k$ and save the tuple $(s_k,
a_k, r_k, s_k')$ as a sample. We also restrict the magnitude of the light and
kilobot movements to get a more realistic simulation.

@@

\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.9\linewidth]{figures/simulator_maze.png}
    \caption{4 kilobots pushing an object through a maze in the simulator.}
    \label{fig:simulator}
\end{figure}

@@

For this policy we decided to only learn how to push the object to the right
instead of an arbitrary direction. This simplifies and speeds up the learning
process because we need to generate fewer samples and can represent our policy
using a smaller state subset for the sparse GP and therefore can compute the
policy update much faster. As discussed in the last paragraph of the previous
section we can still use this policy to push the object in an arbitrary
direction by rotating the state and action appropriately.

@@

We repeat these steps a fixed number of times and manually inspect the learned
policy. If it is not satisfactory (it can not solve the maze) we continue the
learning process.

@@

With $n$ kilobots our state tuple is
$$s = (l_x, l_y, k_{1x}, k_{1y}, \dots, k_{nx}, k_{ny}), $$
where $(l_x, l_y)$ is the light position and $(k_{ix}, k_{iy})$ is the position
of the $i-th$ kilobot. All positions are relative to the object's center
position.

Our action tuple is
$$a = (a_x, a_y),$$
which is the direction the light should move in.

@@

Our reward is given by
$$ r = (c_x - p_x) - 0.5 \cdot |c_y - p_y|, $$
where $(c_x, c_y)$ is the object's current position and $(p_x, p_y)$ is the
object's position from the last time step. Therefore we reward movement to the
right and punish movement to the left and any movement in $y$ direction.

@@

For our learning process we need a feature function $\phi(s, a)$ over
state-action pairs to represent our Q-function, a feature function $\phi(s)$
over states for REPS and a kernel function $K(s_1, s_2)$ between states for the
GP.

@@

After that we compute the distance for the other dimensions (light positions for
$K$ and light positions and actions for $K'$) as
$$
d_{other} = (x_1 - x_2)^T \mathrm{diag}\left(\frac{1}{\sigma_{x}^2}\right) (x_1 - x_2),
$$
where $x_1$ and $x_2$ are vectors holding the values for all non-kilobot
dimensions and $\sigma_x$ is the non-kilobot bandwidth.

@@

In our experiments we used $w = 0.5$ and recomputed $\sigma_{kb}$ and $\sigma_x$
in each iteration as the square root of the median of the squared distances
between all sample values for each dimension separately. We also scaled the
bandwidths uniformly in each dimension by some manually specified factors.

@@

For our final evaluation we made 20 runs of our learning process with 15
iterations each generating $25 \cdot 100 = 2500$ samples per iteration. Using
an Haswell Quad Core with 3.4GHz per core each iteration took about 3 minutes,
so the total run-time for each policy was about 45 minutes.

The policies start of only being able to push the object when the kilobots start
to the left of the object. After about 5 iterations they can also handle kilobot
starting positions above and below the object by first moving the kilobots to
the left and down and then pushing the object from the left side. After further
iterations the policies are able to push the object to the right even if the
kilobots start above and to the right of the object (see
\autoref{fig:policyEvolution}). With even more iterations the kilobot starting
positions from which the object can be pushed tend to extend even further, but
this improvement is inconsistent and the policies may even get worse again.
Therefore we manually inspected the policies at each iteration and chose a good
policy (which is typically found in the first about 15 iterations) for solving
the maze.

\begin{figure*}[!htb]
	\centering
	\subfloat[Policy after 1 iteration]{
        \def\svgwidth{0.45\textwidth}
        \input{figures/policy_1.pdf_tex}
    }
    \hfill
	\subfloat[Policy after 5 iterations]{
        \def\svgwidth{0.45\textwidth}
        \input{figures/policy_5.pdf_tex}
    }
    \hfill
	\subfloat[Policy after 10 iterations]{
        \def\svgwidth{0.45\textwidth}
        \input{figures/policy_10.pdf_tex}
    }
    \hfill
	\subfloat[Policy after 15 iterations]{
        \def\svgwidth{0.45\textwidth}
        \input{figures/policy_15.pdf_tex}
    }
    \hfill
	\caption[foo]{Evolution of the learned policy}
	\label{fig:policyEvolution}
\end{figure*}

The value function (after 8 iterations) also shows that kilobot positions left
of the object are preferable to positions to the right. Further it increases
when moving around the object from the right to the left (see \autoref{fig:value}).

@@

\begin{figure}[!htb]
    \centering
    \resizebox{\linewidth}{!}{\input{figures/reward.tex}}
    \caption{Total reward for each iteration averaged over 20 runs. The shaded
             area shows twice the standard deviation.}
    \label{fig:meanReward}
\end{figure}

We made the policy plots by getting the mean action for states distributed on a
grid and plotting it as an arrow indicating the direction the light should move
in. For this we set the relative light positions to evenly spaced values between
$-0.25$ and $0.25$ for the $x-$ and $y-$dimensions and set the position for all
kilobots to coincide with the light position.

@@

To evaluate our policies we used our simulator to push the object through the
maze as described in \ref{sec:simulator}. We found that our learned policies are
sufficient to achieve this task and are even able to move the object when the
kilobots are (almost) completely on the wrong side of the object, which requires
moving them around the object. Moreover we can transfer our policy to more or
less kilobots and are still able to push the object because of our choice of
kernel. We learned the policy with 4 kilobots and tested it with 1,2,3,4,6,8 and
10 kilobots and it was able to push the object through the maze in each case.
Finally we were also able to push a circle through the maze even though the
policy was learned on a square.

@@

Although this policy works well in the simulator it is not directly applicable
to the real kilobots because our simulator is overly simplistic. It moves
the kilobots directly toward the light instead of using Phototaxis and also
allows for arbitrary light movements, where in reality the light would be moved
by a robot arm which has some constraints on the forces it can apply.

Therefore a future work could simulate the kilobots more realistically and try
to transfer the learned policy to the real kilobots by building a maze and
controlling a robot arm holding a flashlight using the learned policy.
